---
title: "Specificiation and Design of `rdf4r`"
author: "Viktor Senderov"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rdf4r specification and design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

`rdf4r` is an R-library designed to enable the R-practicioner to handle Resource Description Framework (RDF) data. It provides functions for:

- [Interfacing with a triple store](#interface-with-a-triple-store)

## Interface with a triple store

Connect, run queries on, and upload data to a triple store. Functions are separated into [low-level](#low-level-access) access functions that create connections, send raw data, read queries, and [high-level](#high-level-functions) functions that let you create specific functions for your project via factory functions.

### Low-level access

Here's an overview of the low-level access functions:

| Function | Purpose |
|----------|---------|
| [basic_triplestore_access](#connecting-to-a-triplestore) | connecting to a triplestore |
| [api_triplestore_access](#connecting-to-a-triplestore) | connecting to a triplestore |
| [get_protocol_version](#connecting-to-a-triplestore) | protocol version at endpoint |
| [list_repositories](#working-with-repositories) | listing repositories at endpoint |


##### Connecting to a triplestore

In order to access the triplestore, you first need to create a `triplestore_access_options` objects (internally a list). This object can be created by `basic_triplestore_access` in the case of no authentication or basic HTTP authentication or by `api_triplestore_access` in the case of API secret key authentication. They will fail if the URL or the authentication is wrong. The repository information can be omitted at this stage (see [list_repositories](#working-with-repositories)). In case of success the `$status` field of the list will be set to the protocol version. The protocol version can also be queried directly with `get_protocol_version`.

```{r}
# Access with basic HTTP authentication
graphdb = basic_triplestore_access(
  server_url = "http://graph.openbiodiv.net:7777",
  user = "dbuser",
  password = "public-access"
)

# Access with no authentication
graphdb2 = basic_triplestore_access(
  server_url = "http://graph.openbiodiv.net:7777"
)

# Access with API secret key authentication
graphdb3 = api_triplestore_access(
  server_url = "https://rdf.ontotext.com/4135593934/openbiodiv", 
  api_key = "s4bb1d43uc52",
  api_secret = "d7h7eg4e263ghss"
)

# Querying the protocol version for a triplestore endpoint.
get_protocol_version(graphdb)
```

##### Working with repositories

Use `list_repositories` to see what repositories the endpoint offers. The repo id's are given in the column `$id`. You can assign repo id's to the `$repository` field of the `triplestore_access_options` object.

```{r}
list_repositories(graphdb)
graphdb$repository = list_repositories(graphdb)$id[8]
```

##### Submitting data and querying a repository

After you've [selected an endpoint](#connecting-to-a-triplestore) and [a repository](#working-with-repositories), you can add Turtle data to it via `add_data` and submit a SPARQL query via `submit_sparql`. By default results are returned as a dataframe (`as_dataframe = TRUE`). If you override this parameter to `submit_sparql` the results are returned as a raw response that must be parsed (i.e. as JSON).

```{r}
# Submit a SPARQL query
query = "select * where {
  ?s ?p ?o .
  } limit 100"
submit_sparql(query = query, access_options = graphdb)

# Adds Seralized data to store
add_data("PREFIX owl: http://#... \n http://openbiodiv.net/sample_id a owl:Thing.", s)
```


The basic building blocks of RDF are resource identifiers and atomic values.
