---
title: "RDF4R: An R Package for RDF Management.\nProgrammer's Documentation"
author: "Viktor Senderov"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rdf4r specification and design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




RDF4R (`library(rdf4r)`) is an R package designed to enable the
R-practitioner to handle Resource Description Framework (RDF) data. This
is the programmer's documentation, i.e. the "specs and design sheet."





## Table of contents

1. [Interact with a Triple Store](#interact-with-a-triple-store)
2. [Generate RDF](#generate-rdf)






## Interact with a Triple Store

Connect, run queries on, and upload data to a triple store. Functions are separated into [low-level](#low-level-access) access functions that create connections, send raw data, read queries, and [high-level](#high-level-functions) functions that let you create specific functions for your project via factory functions.

Here's an overview of the *low-level access functions*:

[basic_triplestore_access(server_url, user, password, repository)](#connecting-to-a-triplestore) : connecting to a triplestore with no or basic auth 
[api_triplestore_access(server_url, api_key, api_secret, repository)](#connecting-to-a-triplestore) : connecting to a triplestore with a secret key 
[get_protocol_version(access_options)](#connecting-to-a-triplestore) : protocol version at endpoint
[list_repositories(access_options)](#working-with-repositories) : listing repositories at endpoint
[submit_sparql(query, access_options)](#submitting-data-and-querying-a-repository) : executing a SPARQL query [add_data(rdf_data, access_options, data_format)](#submitting-data-and-querying-a-repository) : adding statements (e.g. Turtle) to a repo 
[submit_sparql_update(query, access_options)](): against the update interface

The premise of the high-level functions is to allow the programmer to create their own functions (closures) hiding the database access, the SPARQL code, and other details, and only having parameters of interest.

Here's an overview of the *high-level functions*:

[query_factory(p_query, submit_function access_options, prefixes)](#manufacturing-read-and-UPDATE-query functions): returns a parameterized function executing the query on the endpoint
[add_data_factory(access_options, prefixes)](#Manufacturing-add-data-functions) : Wraps `add_data` to simply submit a Turtle/Trig file to a triplestore 









### Connecting to a triplestore

In order to access the triplestore, you first need to create a `triplestore_access_options` objects (internally a list). This object can be created by `basic_triplestore_access` in the case of no authentication or basic HTTP authentication or by `api_triplestore_access` in the case of API secret key authentication. They will fail if the URL or the authentication is wrong. The repository information can be omitted at this stage (see [list_repositories](#working-with-repositories)). In case of success the `$status` field of the list will be set to the protocol version. The protocol version can also be queried directly with `get_protocol_version`.

*Initializing some connections*:

```{r}
# Access with basic HTTP authentication
graphdb = basic_triplestore_access(
  server_url = "http://graph.openbiodiv.net:7777",
  user = "dbuser",
  password = "public-access"
)

# Access with no authentication
graphdb2 = basic_triplestore_access(
  server_url = "http://graph.openbiodiv.net:7777"
)

# Access with API secret key authentication
#graphdb3 = api_triplestore_access(
#  server_url = "https://rdf.ontotext.com/4135593934/openbiodiv", 
#  api_key = "s4bb1d43uc52",
#  api_secret = "d7h7eg4e263ghss"
#)

# Querying the protocol version for a triplestore endpoint.
get_protocol_version(graphdb)
get_protocol_version(graphdb2)
```








### Selecting a repository

Use `list_repositories` to see what repositories the endpoint offers. The repo id's are given in the column `$id`. You can assign repo id's to the `$repository` field of the `triplestore_access_options` object.

*Selecting a repository*:
```{r}
list_repositories(graphdb)
graphdb$repository = list_repositories(graphdb)$id[8]
```









### Submitting Turtle and executing SPARQL

After you've [selected an endpoint](#connecting-to-a-triplestore) and [a repository](#working-with-repositories), you can add Turtle data to it via `add_data` and submit a SPARQL query via `submit_sparql`. By default results are returned as a dataframe (`as_dataframe = TRUE`). If you override this parameter to `submit_sparql` the results are returned as a raw response that must be parsed (i.e. as JSON).

```{r}
# Adds Seralized data to store
add_data("@prefix owl: <http://www.w3.org/2002/07/owl#>.\n
          @prefix rdfs: <http://www.w3.org/2000/01/rdf-schema#>.\n
          <http://openbiodiv.net/example1> a owl:Thing;\n
                                    rdfs:label 'eg1'.",
         access_options = graphdb)

# Submit a SPARQL query
query = "select * where {
  <http://openbiodiv.net/example1> ?p ?o .
  } limit 100"
submit_sparql(query = query, access_options = graphdb)
```










### Manufacturing query functions

The function `query_factory`'s purpose is to manufacture a function executing a specified SPARQL query against a specified endpoint. The SPARQL query that the function takes as an input

```{r query_factory, results = "hide"}
p_query = "SELECT DISTINCT ?id WHERE {
  ?id rdfs:label %label
}"

p_query2 = "SELECT * WHERE {
  ?s ?p ?o
} LIMIT 5"

drop_query = "DROP GRAPH %subgraph"

simple_lookup = query_factory(p_query, access_options = graphdb)
simplest_f = query_factory(p_query2, access_options = graphdb)
drop_g = query_factory(drop_query, submit_function = submit_sparql_update, access_options = graphdb_secret)

simplest_f()
simple_lookup("'eg1'")
#simple_lookup(label = "Pavel Stoev")
#drop_g("<http://openbiodiv.net/123>")
```

Note that the queries are specified in a *parameterized format*. I.e. the strings indicated by `%<something>` will get replaced by the parameters of the manufactured function. There can be a query with no parameters. If you wish to see what are the arguments of a manufactured function later (as it has no manpage), you can use `args`. E.g.:

```{r}
args(simple_lookup)
```






### Manufacturing unctions that add data

```{r}
prefixes = c(rdfs = "<http://www.w3.org/2000/01/rdf-schema#>", foaf = "<http://xmlns.com/foaf/0.1/>", openbiodiv = "<http://openbiodiv.net/>")
add_data_to_graphdb = add_data_factory(access_options = graphdb, prefixes = prefixes)


ttl = "openbiodiv:examples {
openbiodiv:example2 rdfs:label 'Second example'@en .
}"
add_data_to_graphdb(ttl)

simple_lookup("'Second example'@en")
```








## RDF

The basic building blocks of RDF are resource identifiers and atomic values
(literals). Identifiers and literals are stored as lists with certain 
fields.

[identifier(id, prefix = NA)](#identifier-consturction): constructs an 
  unique identifier from a local id and a prefix.
[fidentifier(label, fun = list(...), prefixes, def_prefix, ...)](#identifier-consturction): construct identifier via a lookup function.
[identifier_factory(fun = list(...), prefixes, def_prefix, ...)](#manufacturing-identifier-constructors)
[literal(text_value, xsd_type, lang)](#literals) constructs a `literal` object
[RDF$new()](#RDF) constructs an `RDF` object









### Literals

To initiliaze literals, you use the `literal` constructor; however, as the
`literal` constructor has a *xsd type* field, you should use one of the
already predefined `?xsd_types`.

```{r}
lking_lear = literal(
  text_value = "King Lear",
  lang = "en"
)

las_you_like_it = literal(
  text_value = "As You Like It",
  lang = "en"
)

lhamlet = literal(
  text_value = "Hamlet",
  lang = "en"
)

lothello = literal(
  text_value = "Othello",
  lang = "en"
)

lsonnet_78 = literal(
  text_value = "Sonnet 78",
  lang = "en"
)

lastrophil = literal(
  text_value = "Astrophil and Stella",
  lang = "en"
)

ledward2 = literal(
  text_value = "Edward II",
  lang = "en"
)

lhero = literal(
  text_value = "Hero and Leander",
  lang = "en"
)

lgreensleeves = literal(
  text_value = "Greensleeves",
  lang = "en"
)

lshakespeare = literal(
  text_value = "Shakespeare"
)

lsir_phillip_sidney = literal(
  text_value = "Sir Phillip Sidney"
)

lchristopher_marlowe = literal(
  text_value = "Christopher Marlowe"
)

lhenry_8_rex = literal(
  text_value = "Henry VII Rex"
)

l1599 = literal(
  text_value = "1599",
  xsd_type = xsd_integer
)

l1603 = literal(
  text_value = "1603",
  xsd_type = xsd_integer
)

l1609 = literal(
  text_value = "1609",
  xsd_type = xsd_integer
)

l1590 = literal(
  text_value = "1609",
  xsd_type = xsd_integer
)

l1592 = literal(
  text_value = "1592",
  xsd_type = xsd_integer
)

l1593 = literal(
  text_value = "1593",
  xsd_type = xsd_integer
)

l1525 = literal(
  text_value = "1593",
  xsd_type = xsd_integer
)
```







### Constructing identifiers

An identifier in the semantic web is something that uniquely identifies a
resource. Identifiers can be represented as URI's (e.g.
<http://example.com/id>), or as QNAME's (e.g. example:id). To consturct
an identifier, you can use one of the two constructor functions:

`identifier` is the basic constructor function. It takes a prefix and a 
local part and returns the object. If multiple prefixes or ids are
supplied, `identifier` will only use the first one.

`fidentifier` is a constructor function that takes a lookup function, or a
list of lookup functions, a label (to be supplied to the lookup function)
and returns the unique identifier it found via the lookup functions. If it
fails, it calls `identifier` with a default prefix and a UUID.

```{r}
play = identifier(
   id = "Play",
   prefix = c(openbiodiv = "http://openbiodiv.net/")
)

poem = identifier(
   id = "Poem",
   prefix = c(openbiodiv = "http://openbiodiv.net/")
)

song = identifier(
   id = "Song",
   prefix = c(openbiodiv = "http://openbiodiv.net/")
)

wrote = identifier(
   id = "wrote",
   prefix = c(openbiodiv = "http://openbiodiv.net/")
)

has_year = identifier(
  id = "has_year",
  prefix = c(openbiodiv = "http://openbiodiv.net/")
)
```

```{r}
another_example = identifier("http://openbiodiv.net/as_you_like")
another_example
```

```{r}
 prefixes = c(
   rdfs = "http://www.w3.org/2000/01/rdf-schema#",
   foaf = "http://xmlns.com/foaf/0.1/",
   openbiodiv = "http://openbiodiv.net/"
 )

fidentifier(
 label = lking_lear$squote,
 prefixes = prefixes,
 def_prefix = c(openbiodiv = "http://openbiodiv.net/"),
 simple_lookup)

fidentifier(
 label = las_you_like_it$squote,
 prefixes = prefixes,
 def_prefix = c(openbiodiv = "http://openbiodiv.net/"),
 simple_lookup)
```

Note that at this point the lookup may fail (and create new identifiers
every time) as we've not submitted our data.






### Manufactureing identifier constructors

`identifier_factory` returns a function with just one argument 
(that has to be of literal type) that wraps around an `findentifier` call.

```{r}
openbiodiv_id = identifier_factory(simple_lookup,
   prefixes = prefixes,
   def_prefix = c(openbiodiv = "http://openbiodiv.net/"))

 idking_lear = openbiodiv_id(lking_lear)
 idas_you_like_it = openbiodiv_id(las_you_like_it)
 idhamlet = openbiodiv_id(lhamlet)
 idothello = openbiodiv_id(lothello)
 idsonnet78 = openbiodiv_id(lsonnet_78)
 idastrophil = openbiodiv_id(lastrophil)
 idedward2 = openbiodiv_id(ledward2)
 idhero = openbiodiv_id(lhero)
 idgreensleeves = openbiodiv_id(lgreensleeves)
 idshakespeare = openbiodiv_id(lshakespeare)
 idsir_phillip_sidney = openbiodiv_id(lsir_phillip_sidney)
 idchristopher_marlowe = openbiodiv_id(lchristopher_marlowe)
 idlhenry_8_rex = openbiodiv_id(lhenry_8_rex)
```















### Triples

An RDF object is a list of triples.

RDF object are unlike literals or identifiers, which are lists and,
therefore immutable. RDF objects are R6 mutable objects. This mutability
implies that one function can create a new RDF object and pass it around,
while other functions grow it.

```{r}

classics_rdf = RDF$new()

classics_rdf$add_triple(
  subject = idshakespeare, predicate = wrote, object = idking_lear
)

classics_rdf$add_triple(
  subject = idking_lear, predicate = rdfs_label, object = lking_lear
)

classics_rdf$add_triple(
  subject = idshakespeare, predicate = wrote, object = idas_you_like_it
)

classics_rdf$add_triple(
  subject = idas_you_like_it, predicate = rdfs_label, object = las_you_like_it
)

classics_rdf$add_triple(
  subject = idas_you_like_it, predicate = has_year, object = l1599
)

classics_rdf$add_triple(
  subject = idas_you_like_it, predicate = rdf_type, object = play
)

classics_rdf$serialize("<http://example.gov/>")
```
