---
title: "Comparison betwen `rdflib` and RDF4R"
output: html_notebook
---

```{r}
library(magrittr)
library(rdflib)
library(rdf4r)
```

## Task

Use both `rdflib` and RDF4R to model Table 3-1 from
[Semantic Web for the Working Ontologist](http://www.kevenlw.name/downloads/Ontologist.pdf).

To make things more interesting use the prefix <http://rdflib-rdf4r.net/> for instances and the finctional art ontology, <http://art-ontology.net/>, to reify the types of works of art.

E.g. for Shakespeare, use <http://rdflib-rdf4r.net/Shakespeare> and for Poem use <http://art-ontology.net/Poem>.

(a) Create an RDF representation of the data.
(b) Serialize the RDF representation.
(c) Submit the serialization to a triple store.
(d) Query the triple store.

## Solution

### (a) with RDF4R

(1) We create the literals, which are special type of list, containing the type, language, and semantic quotation (e.g. "1994"^^xsd:year) that can be used for serialization:

```{r}
lking_lear      = literal(text_value = "King Lear",        lang = "en")
las_you_like_it = literal(text_value = "As You Like It",   lang = "en")
lhamlet         = literal(text_value = "Hamlet",           lang = "en")
lothello        = literal(text_value = "Othello",          lang = "en")
lsonnet_78      = literal(text_value = "Sonnet 78",        lang = "en")
lastrophil      = literal(text_value = "Astrophil and Stella",
                                                           lang = "en")
ledward2        = literal(text_value = "Edward II",        lang = "en")
lhero           = literal(text_value = "Hero and Leander", lang = "en")
lgreensleeves   = literal(text_value = "Greensleeves",     lang = "en")

lshakespeare         = literal(text_value = "Shakespeare")
lsir_phillip_sidney  = literal(text_value = "Sir Phillip Sidney")
lchristopher_marlowe = literal(text_value = "Christopher Marlowe")
lhenry_8_rex         = literal(text_value = "Henry VII Rex")

l1599 = literal(text_value = "1599", xsd_type = xsd_integer)
l1603 = literal(text_value = "1603", xsd_type = xsd_integer)
l1609 = literal(text_value = "1609", xsd_type = xsd_integer)
l1590 = literal(text_value = "1590", xsd_type = xsd_integer)
l1592 = literal(text_value = "1592", xsd_type = xsd_integer)
l1593 = literal(text_value = "1593", xsd_type = xsd_integer)
l1525 = literal(text_value = "1525", xsd_type = xsd_integer)
```

Note that for the work titles, we use the argument `lang="en"` telling the literal constructor that the literal value is in English, whereas for the names, we omit this argument. As per semantic web conventions, when the argument is omitted, and no type is explicitly specified, it is assumed that the literal is a string (`xsd:string`). For the literals containing years, on the other hand, we explicitly specify an integer type; otherwise they would have parsed as strings as well. All of this can be seen by inspecting the individual lists (objects of class `literal` are lists):

```{r}
lhamlet
cat("\n================\n\n")
lshakespeare
cat("\n================\n\n")
l1603
```

(2) We need resource identifiers for our resources, i.e. playwrights, works of art, as well for the classes of which those resources are instances of. To make things simpler, we use a fictional "Art Ontology" with the prefix <http://art-ontology.net/>. Let's "hardcode" identifiers for the ontology classes:

```{r}
prefixes = c(
   rdfs = "http://www.w3.org/2000/01/rdf-schema#",
   rdf = "http://www.w3.org/1999/02/22-rdf-syntax-ns#",
   example = "http://rdflib-rdf4r.net/",
   art = "http://art-ontology.net/"
 )
eg = prefixes[3]
art = prefixes[4]
artist = identifier(id = "Artist", prefix = art)
play = identifier(id = "Play", prefix = art)
poem = identifier(id = "Poem", prefix = art)
song = identifier(id = "Song", prefix = art)
wrote = identifier(id = "wrote", prefix = art)
has_year = identifier(id = "has_year", prefix = art)
```

Let's inspect one class and one property:

```{r}
artist
cat("\n================\n\n")
wrote
```

Note that each `identifier` object is a list where the field `$uri` gives the URI of the resource and the field `qname` gives the shortened name (QNAME) with respect to the prefix stored in `$prefix`.

We also need resource identifiers for our entitites such as Shakespaere, Christopher Whatchamacallit, etc. Usually, we don't want to mint identifiers for entities according to Semantic Web custom. We want to look them up in a database, and only mint if they are not found. For this, RDF4R offers factory functions to create lookup/mint functions. For this example, we have fired up a publicly accessible triplestore running GraphDB and we use the publicly accessible repository `obkms_i7`.

```{r}
graphdb = basic_triplestore_access(
  server_url = "http://graph.openbiodiv.net:7777",
  user = "dbuser",
  password = "public-access"
)
graphdb$repository = "obkms_i7"

p_query = "SELECT DISTINCT ?id WHERE {
  ?id rdfs:label %label
}"

simple_lookup = query_factory(p_query, access_options = graphdb)

lookup_or_mint_id = identifier_factory(simple_lookup,
   prefixes = prefixes,
   def_prefix = eg)

idking_lear = lookup_or_mint_id(lking_lear)
idas_you_like_it = lookup_or_mint_id(las_you_like_it)
idhamlet = lookup_or_mint_id(lhamlet)
idothello = lookup_or_mint_id(lothello)
idsonnet78 = lookup_or_mint_id(lsonnet_78)
idastrophil = lookup_or_mint_id(lastrophil)
idedward2 = lookup_or_mint_id(ledward2)
idhero = lookup_or_mint_id(lhero)
idgreensleeves = lookup_or_mint_id(lgreensleeves)
idshakespeare = lookup_or_mint_id(lshakespeare)
idsir_phillip_sidney = lookup_or_mint_id(lsir_phillip_sidney)
idchristopher_marlowe = lookup_or_mint_id(lchristopher_marlowe)
idlhenry_8_rex = lookup_or_mint_id(lhenry_8_rex)
```

Let's inspect

```{r}
idshakespeare
```

Note that the "mint" functionality has been enabled as the identifiers are not present in the triplestore. Once we submit them later on the `lookup_or_mint` function will return the same identifier everytime we call it with a particular label/literal.

Now to create the RDF representation:

```{r chunk, results = "hide"}
classics_rdf = ResourceDescriptionFramework$new()
classics_rdf$add_triple(subject = idshakespeare,    predicate = wrote,      object = idking_lear)
classics_rdf$add_triple(subject = idking_lear,      predicate = rdfs_label, object = lking_lear)
classics_rdf$add_triple(subject = idshakespeare,    predicate = wrote,      object = idas_you_like_it)
classics_rdf$add_triple(subject = idas_you_like_it, predicate = rdfs_label, object = las_you_like_it)
classics_rdf$add_triple(subject = idas_you_like_it, predicate = has_year,   object = l1599)
classics_rdf$add_triple(subject = idas_you_like_it, predicate = rdf_type,   object = play)
```

The easiest way to inspect the `ResourceDescriptionFramework` object is to actually serialize it. The serialization function takes one argument (the subgraph where the triples should be stored). We will reuse the example for that.

```{r}
cat(classics_rdf$serialize(strip_angle(eg, reverse = TRUE)))
```

### (a) with `rdflib`

```{r}
classics_rdflib <- rdf()
rdf_add(classics_rdflib,
             subject   = strip_angle(idshakespeare$uri),
             predicate = strip_angle(wrote$uri),
             object    = strip_angle(idking_lear$uri)
        )

rdf_add(classics_rdflib,
        subject   = strip_angle(idshakespeare$uri),
        predicate = strip_angle(wrote$uri),
        object    = strip_angle(idas_you_like_it$uri)
      )

rdf_add(classics_rdflib,
        subject   = strip_angle(idas_you_like_it$uri),
        predicate = strip_angle(rdfs_label$uri),
        object    = las_you_like_it$text_value,
        datatype_uri = strip_angle(las_you_like_it$xsd_type$uri)
)

rdf_add(classics_rdflib,
        subject   = strip_angle(idas_you_like_it$uri),
        predicate = strip_angle(has_year$uri),
        object    = l1599$text_value,
        datatype_uri = strip_angle(l1599$xsd_type$uri)
      )

rdf_add(classics_rdflib,
        subject   = strip_angle(idas_you_like_it$uri),
        predicate = strip_angle(rdf_type$uri),
        object    = strip_angle(play$uri)
)

out <- tempfile("file", fileext = ".rdf")
rdf_serialize(classics_rdflib, doc = out, format = "turtle", namespace = prefixes)
cat(do.call(paste, c(as.list(readLines(out)), list(sep = "\n"))))
```
